---
title: "PA1_template.Rmd"
author: "Clement Chong"
date: "30/07/2020"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(zip)
library(dplyr)
library(ggplot2)
```

We first download and unzip the file and then read it into a dataframe in R:

```{r}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile = "data.zip", mode = "wb")
unzip("data.zip")
activity <- read.csv("activity.csv")
```

The next step is to calculate all of the value we need for the plot:

```{r}
days <- group_by(activity,date) #creating group by date

#finding total steps per day
total.steps <- summarise(days,sum(steps, na.rm = TRUE))
names(total.steps) <- c("date","steps")
avgsteps <- mean(total.steps$steps) #mean
medsteps <- median(total.steps$steps) #median
```

The results show that the mean total number of steps taken a day is 'r avgsteps' and the median is 'r medsteps' and the plot is as below:

```{r}
hist(total.steps$steps, main = "Histogram of Total Steps Taken per Day", xlab = "Total Steps per Day")
```

next, we will calculate the average number of steps for each interval across all days and plot it:

```{r}
minint <- group_by(activity, interval)
dayavg <- summarise(minint,mean(steps, na.rm = TRUE))
names(dayavg) <- c("interval","steps")
plot(dayavg$interval,dayavg$steps,type = 'l', main = "Time Series of average Steps Each Interval", xlab = "Interval", ylab = "Average Steps Over the Days")

maxavgsteps <- dayavg$interval[which.max(dayavg$steps)] 
```

as you can see from the plot, the average maximum steps occurs at interval `r maxavgsteps`

```{r}
missing <- sum(!complete.cases(activity))
```

and the total number of missing cases is `r missing`

Now we will make a new dataset which is equal to the existing one but with the missing values filled in with the average steps for that interval:

```{r}
newact <- activity
naind <- which(is.na(activity$steps))
  for (i in 1:length(naind))
  {
    indx <- naind[i]
    newact$steps[indx] <- dayavg$steps[dayavg$interval == newact$interval[indx]]
  }
newdays <- group_by(newact,date) 
newtotal.steps <- summarise(newdays,sum(steps, na.rm = TRUE))
names(newtotal.steps) <- c("date","steps")
newavgsteps <- mean(newtotal.steps$steps)
newmedsteps <- median(newtotal.steps$steps)

hist(newtotal.steps$steps, main = "Histogram of Total Steps Taken per Day", xlab = "Total Steps per Day")
```

As seen from the histogram above, the mean total number of steps taken a day is 'r newavgsteps' and the median is 'r newmedsteps`. The histogram also shows that with the newly filled in data, the plot takes on a more normal distribution with most values in the middle and less towards the extremes.

Next up, we will investigate if there are differences in activity patterns between weekdays and weekends.

```{r}
daysoftheweek <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
newweek <- mutate(newact, weekday = factor((weekdays(as.Date(activity$date)) %in% daysoftheweek), levels=c(FALSE, TRUE), labels=c('weekend', 'weekday')))

lastdata <- group_by(newweek,weekday,interval)
weeksteps <- summarise(lastdata,mean(steps))
names(weeksteps) <- c("weekday","interval","steps")

ggplot(weeksteps, aes(x = interval, y = steps)) +
      geom_line() +
      facet_grid(weekday ~ .)

```

The plot shows that the activity patterns are similar on weekdays and weekends but much more movement is recorded between the intervals of 500 and 800 on weekdays as well as a higher peak at about 800. However, movement after 1000 and before 2100 is slightly higher for weekends.
